---
title: "Meta-analysis vs. Bayesian Evidence Synthesis"
author: "Elise van Wonderen"
date: "21 January 2022"
output: 
 ioslides_presentation:
  toc: true
  logo: UU.png
  smaller: TRUE
---

```{r, echo=F, message=F, warning=F}
library(plotly)
library(DT)
library(tidyverse)
source("functions.R")
```

## Content

- Introduction
    - Research synthesis
    - Comparison of meta-analysis and BES
    - Meta-analysis
    - Bayesian evidence synthesis
- Simulation
    - Data generation
    - Sampling population effect sizes
    - Hypotheses
    - Simulation results I
    - Simulation results II


# Introduction

## Research synthesis

- Many conflicting results in psychological science (OSC, 2015) 
<br>
<br>
- Further exacerbated by small sample sizes and reliance on statistical significance tests (Button et al., 2013; van Calster et al., 2018) 
<br>
<br>
- Increasing the robustness of results by quantitatively summarizing multiple studies
    - Meta-analysis
    - Bayesian Evidence Synthesis (BES)
    
    
## Comparison of meta-analysis and BES

<br>

<div style="float: left; width: 48%;">
**META-ANALYSIS** <br>
<br>
Combines effect sizes <br>
<br>
<br>
Asks what the average effect is in all studies combined (data-pooling method) <br>
<br>
Effect sizes need to have the same metric and come from the same model
</div>

<div style="float: right; width: 48%;">
**BAYESIAN EVIDENCE SYNTHESIS** <br>
<br>
Combines support for a specific hypothesis <br>
<br>
Asks to what extent a hypothesis is supported in *each* study (NOT a data-pooling method) <br>
<br>
There are no restrictions on which estimates can be combined
</div>



## Meta-analysis

Computing meta-analytic effect size $\bar{\beta}$ and its variance $v$ (Hedges & Vevea 1998):

$$ 
\begin{aligned}
\bar{\beta} &= \frac{\sum_{k=1}^{K} w_k\hat{\beta_k}}{\sum_{k=1}^{K} w_k} \\
\\
w_k &= \frac{1}{var(\hat{\beta_k})} \\
\\
v &= \frac{1}{\sum_{k=1}^{K} w_k}
\end{aligned}
$$
<br>
where $K$ is the number of studies, and $\hat{\beta_k}$ is the estimated effect size of study $k$.



## Bayesian evidence synthesis

- Combine studies at the level of the hypothesis instead of the effect size
- Compute study-specific Bayes factors and turn these into posterior model probabilities (PMPs)
- Then compute an aggregated PMP over all studies (Kuiper et al., 2012):

<br>

$$
\pi_{K,h} = \frac{\prod_{k=1}^{K}\pi_{k,h}}{\sum_{h=1}^{H}\prod_{k=1}^{K}\pi_{k,h}}
$$
<br>
where $h$ is hypothesis 1, ..., $H$ and $\pi$ is the PMP.


# Simulation

## Data generation

Generate data from a regression model with one predictor:
$$ Y = \beta X_1 + \epsilon, $$
where $X_1$ is sampled from $\mathcal{N}(0,1)$ and $\epsilon$ is sampled from from $\mathcal{N}(0,\sqrt{1-R^2})$.

<br>

```{r, eval=F}
beta <- 0.3                                                    # specify beta value
R2 <- beta^2                                                   # R-squared
n <- 25                                                        # sample size
X <- rnorm(n = n, mean = 0, sd = 1)                            # sample X values
Y <- X * beta + rnorm(n = n, mean = 0, sd = sqrt(1 - R2))      # compute Y
```


## Sampling population effect sizes

Instead of assuming one underlying population effect size, we assume study-specific population effect sizes that randomly deviate from the overall mean following a normal distribution $\mathcal{N}(0.3,0.1)$:

```{r histogram, echo=F, cache=T, message=F, warning=F, fig.align='center', fig.height=3.9, fig.width=6}
rnorm(n = 2000, mean = 0.3, sd = 0.1) %>% 
  as.data.frame() %>% 
  ggplot() +
  geom_histogram(aes(.), fill = "lightgrey", col = "black") +
  xlim(c(-0.2, .70)) +
  geom_vline(xintercept = 0.3, lty=2, col = "blue", size=1) +
  theme_classic()  +
  theme(axis.text = element_text(size=12),
        axis.title = element_text(size=12))
```

## Hypotheses

We tested $H1: \beta > 0$ against either the null hypothesis ($H0$), the complement hypothesis ($Hc$) or the unconstrained hypothesis ($Hu$):

$$
\begin{aligned}
H0&: \ \beta = 0 \\
Hc&: \ \beta < 0 \\
Hu&: \ \beta
\end{aligned}
$$  
where $Hu$ means that $\beta$ can take on any value.


## Simulation results I


```{r datatable, echo=F, warning=F, message=F, cache=T}
dat <- readRDS("01_equal_sample_sizes.rds")
dat %>% 
  filter(nsim == 1) %>% 
  dplyr::select(-data) %>% 
  round(digits = 2) %>% 
  datatable(options = list(pageLength = 5))
```

## Simulation results II

Proportion of simulations (nsim = 1000) that H1 gets selected over the alternative hypothesis based on the aggregated PMP:

```{r simulation plot, echo=F, warning=F, message=F, fig.align='center', fig.height=3.9, cache=T}
p <- dat %>% 
  group_by(n, nsim) %>%
  mutate(TH10 = best_hypothesis(PMP10),
         TH1c = best_hypothesis(PMP1c),
         TH1u = best_hypothesis(PMP1u)) %>%  
  dplyr::filter(n_study %in%  c(2, 4, 8)) %>%
  mutate(n_study = as.factor(n_study)) %>%
  group_by(n, n_study) %>%
  summarize(H10 = sum(TH10)/n(),
            H1c = sum(TH1c)/n(),
            H1u = sum(TH1u)/n()) %>%
  gather(key = "method", value = "THR",
         H10, H1c, H1u,
         factor_key = T) %>%
  ggplot(mapping = aes(x = n, y = THR, col = n_study, lty= n_study, shape = n_study)) +
  scale_x_continuous(breaks = seq(15, 120, 45), labels=seq(15, 120, 45))  +
  geom_line(size =1) +
  geom_point(size =2) +
  geom_hline(yintercept = 0.9, lty=2, col="lightgray") +
  facet_wrap(~method, nrow = 1) +
  labs(x = "Sample size", y = "") +
  theme_classic() +
  theme(axis.text = element_text(size =10))

ggplotly(p)
```




## References

Button, K. S., Ioannidis, J. P., Mokrysz, C., Nosek, B. A., Flint, J., Robinson, E. S., & Munafò, M. R. (2013). Power failure: why small sample size undermines the reliability of neuroscience. *Nature Reviews Neuroscience, 14*(5), 365-376. https://doi.org/10.1038/nrn3502 

Hedges, L. V., & Vevea, J. L. (1998). Fixed-and random-effects models in meta-analysis. *Psychological Methods, 3*(4), 486. https://doi.org/10.1037/1082-989X.3.4.486

Kuiper, R. M., Buskens, V., Raub, W., & Hoijtink, H. (2013). Combining statistical evidence from several studies: A method using Bayesian updating and an example from research on trust problems in social and economic exchange. *Sociological Methods & Research, 42*(1), 60-81. https://doi.org/10.1177/0049124112464867

Open Science Collaboration (2015). Estimating the reproducibility of psychological science. *Science, 349*(6251), aac4716. https://doi.org/10.1126/science.aac4716 

Van Calster, B., Steyerberg, E. W., Collins, G. S., & Smits, T. (2018). Consequences of relying on statistical significance: Some illustrations. *European Journal of Clinical Investigation, 48*(5), e12912.  https://doi.org/10.1111/eci.12912  








